### 2023-09-20 main.tex
check partial

\newcommand{\changelocaltocdepth}[1]{%
  \addtocontents{toc}{\protect\setcounter{tocdepth}{#1}}%
  \setcounter{tocdepth}{#1}%
}
\newcommand{\id}{I}
\newcommand{\Xt}{\tilde{X}}

@@

@@

\subsection{Duality for the integrable process on a line segment}
Here we consider the chain defined in Subsection \ref{subsection-description-process-LINE}. In this setting the extra-sites are denoted by $0$ and $L+1$, connected respectively with sites $1$ and $L$ of the chain. Moreover, the matrix $R_{x}$ defined in \eqref{Revmatrix} 
reduces to the identity which implies 

@@

Therefore, the duality matrix reads

@@

and the dual Hamiltonian 
	\widetilde{H}_{left}=|\alpha|\sum_{a=1}^{N-1}\left((a^{\dagger})_{a}^{0}\,E_{Na}^{1}-E_{aa}^{1}\right)\qquad 	\widetilde{H}_{right}=|\alpha|\sum_{a=1}^{N-1}\left((a^{\dagger})_{a}^{L+1}\,E_{Na}^{L}-E_{aa}^{L}\right)
In the long-time limit, the dual process voids the chain, and all particles are eventually absorbed at the extra-sites $0$ and $L+1$.\\
In this situation the elements of the duality matrix \eqref{dualityElements} read

@@

Using the above duality elements, we can compute the  $m$-point correlations between non-empty particles using the so called absorption probabilities. Consider a dual configuration $\hat{\xi}:=\sum_{k=1}^{m}\delta_{a_{k}}^{x_{k}}$ where $a_{k}\in\{1,\ldots,N-1\}$ and $x_{k}\in\{1,\ldots,L\}$ for every $k\in\{1,\ldots,m\}$. Considering an arbitrary $\bm{n}$, the elements \eqref{ElementsDualityMatrixChain} of the duality matrix computed in the dual configuration $\hat{\xi}$ read
	D(\bm{n},\hat{\xi})=\prod_{k=1}^{m}\mathbbm{1}_{\{n_{a_{k}}^{x_{k}}=1\}}
To compute the steady state $m$-point correlations of the process $\bm{n}(t)$ we recall that the dual process eventually voids the chain (by putting all particles in the extra-sites $0$ and $L$). Therefore, we have the following equalities
\begin{equation}\label{CorrelationByABS-prob}
	\langle n_{a_{1}}^{x_{1}}\cdots n_{a_{m}}^{x_{m}}\rangle_{stat}&=\mathbb{E}_{stat}\left[D(\bm{n},\hat{\xi})\right]=\lim_{t\to \infty}\mathbb{E}_{\bm{n}}\left[D(\bm{n}(t),\hat{\xi})\right]=\lim_{t\to \infty}\mathbb{E}_{\bm{\xi}}\left[D(\bm{n},\hat{\xi}(t))\right]
	\\&=\langle\hat{\bm{\xi}}\rangle_{stat,dual}=\sum_{c_{1}=0}^{1}\ldots \sum_{c_{m}=0}^{1}\prod_{k=1}^{m}\alpha_{a_{k}}^{c_{k}}\beta_{a_{k}}^{1-c_{k}}\mathcal{P}(c_{1},\ldots,c_{m},\hat{\bm{\xi}})
\begin{equation}\label{absProbabilitiesIntegrable}
	\mathcal{P}(c_{1},\ldots,c_{m},\hat{\bm{\xi}}):=\mathbb{P}\left(\bm{\xi}(\infty)=\sum_{k=1}^{m}\left(c_{a}\delta_{0}+(1-c_{a})\delta_{L+1}\right)| \bm{\xi}(0)=\hat{\bm{\xi}}\right)
and they are called \textit{absorption probabilities}, namely the joint probability of having, for all $k=1,\ldots,m$, $c_{k}$ particles of species $a_{k}$ absorbed at left and $1-c_{k}$ absorbed at right. Therefore, by computing these absorption probabilities we determine the arbitrary non-equilibrium steady state correlations of the integrable multi-species stirring process.

@@

In this section we aim to use duality and the matrix product ansatz to determine the non equilibrium steady state and its arbitrary correlation for the integrable version of the multi-species stirring process, defined in Subsection \ref{subsection-description-process-LINE}. As it will turn out, duality simplifies the commutation relations of the matrix product ansatz allowing to determined a closed formula for all correlations. As a by-product, absorption probabilities can be explicitly determined, giving a further probabilistic interpretation for the non-equilibrium steady state. 

@@

This Markov chain is positive recurrent, therefore there exists a unique stationary measure, that will be reached when time goes to infinity, regardless of the initial distribution. 

@@

The MPA gives an abstract form of the steady state in terms of an algebra of operators. However, to write explicitly the steady state one should provide a representation of this algebra. Another approach is to directly use the commutators and and the boundary actions compute the steady-state correlations. This path has been followed in \cite{vanicat2017exact} to compute one-point and two-point correlations. 
Here we follow the same approach. We will show that the duality property established in Section \ref{sectionDuality} simplifies substantially the  use of commutators in the actual

@@

The first transformation $S_{1}$ is closely related to the dual matrix. Moreover, we will see that $H^{'}$ is $\widetilde{H}^{T}$, up to the extra-site term described by the bosonic creation operator in \eqref{siteDualHamiltonian}. Therefore, $H^{'}$ is not stochastic. The second transformation $S_{2}$ diagonalize the right boundary. The composition  $S_{2} S_{1}$ allows to simplify the commutators of the MPA algebra. \\

The computation are organized as follows: we first define the similarity transformations $S_{1},S_{2}$ and the Hamiltonians $H^{'},H^{''}$. Second, we compute right eigenvector with vanishing eigenvalue of $H^{''}$ thanks to the simpler commutators. Finally, we invert the similarities and retrieve the right eigenvector with vanishing eigenvalue of $H^{'}$ and  the arbitrary correlations and the absorption probabilities. The explicit expression of the steady state of $H$ is reported in appendix \ref{appB}.

@@

\textbf{Remark:} the boundary Hamiltonians $H_{left}^{'}$ and $H_{right}^{'}$ relates to the transposed of  $\widetilde{H}_{left}$ and $\widetilde{H}_{right}$, the boundary part of the dual Hamiltonian  defined in \eqref{siteDualHamiltonian}. They are obtained replacing the  extra-site bosonic creation operators $(a^{\dagger})_{a}^{0}$  (resp. $(a^{\dagger})_{a}^{L+1}$) with the corresponding reservoir parameters $\alpha_a$ (resp. $\beta_a$).

@@

i.e. the left one is lower triangular and depends on the differences between boundary parameters, while the right one is diagonal. The expressions of $H_{left}^{''}$ and $H_{right}^{''}$ are proved using the Hadamard formula \eqref{HadamardFormula} with $B=\sum_{c=1}^{N-1}\beta_{c}e_{c1}$ and $A=e_{aN},$ and $A=e_{aa}$. We compute

@@

It is convenient to introduce also the transformation

@@

    \delta_{\tau_{x},1}:=\begin{cases}

@@

Using the transformed operators $(\widetilde{X}_{a})_{a\in\{1,\ldots,N\}}$, the vector $|\Psi^{''}\rangle$ is written as:

@@

Knowing the steady state  of the Hamiltonian $H^{''}$ we use \eqref{S2-Inverse} to retrieve the steady state of the Hamiltonian $H^{'}$. The result is the following 

@@

We consider a chain of length $L$ where the end sites $1$ and $L$ are connected with two reservoirs. Now, each site can host $\nu\geq 1$ particles, chosen among the $N$ types. The process $\left(\bm{n}(t)\right)_{t\geq0}$ is defined on the space $\Omega=\bigtimes_{x=1}^{L} \Omega_{x}$ and the generator reads
This process is dual with respect to $(\bm{\xi}(t))_{t\geq0} $ defined on the enlarged configuration space
	\widetilde{\Omega}=N_{0}^{N-1}\times\bigtimes_{x=1}^{L} \Omega_{x}\times N_{0}^{N-1}
where the boundaries are absorbing, i.e. 
		\widetilde{\mathcal{L}}_{left}f(\bm{\xi})&=|\alpha|\sum_{a=1}^{N-1}\xi_{a}^{1}\left(f(\bm{\xi}-\bm{\delta}_{a}^{1}+\bm{\delta}_{a}^{0}+\bm{\delta}_{a}^{1})f(\bm{\xi})\right)\\\widetilde{\mathcal{L}}_{right}f(\bm{\xi})&=|\beta|\sum_{a=1}^{N-1}\xi_{a}^{L}\left(f(\bm{\xi}-\bm{\delta}_{a}^{L}+\bm{\delta}_{a}^{L+1}+\bm{\delta}_{a}^{L})-f(\bm{\xi})\right)
The dual process has absorbing boundaries, i.e. all dual particles are eventually absorbed at $0$ or at $L+1$ with a certain probability. This implies that, the chain is eventually voided and the steady state correlation of the original process can be written via the absorption probabilities of the dual process as explained in the following. 
We call $\mu_{stat}$ the stationary measure of the the process $(\bm{n}(t))_{t\geq 0}$. We assume that it is initially distribute with a configuration $\bm{n}$. For every $m\in\{1,\ldots,L\}$ we consider the dual configuration  $\bm{\widehat{\xi}}=\delta_{\tau_{1}}^{y_{1}}+\ldots+\delta_{s_{m}}^{y_{m}}$ with $y_{x}<y_{x+1}$  and $s_{x}\in \{1,\ldots,N-1\}$ for every $x\in \{1,\ldots,m\}$. The following chain of equalities holds
\mathcal{P}(i_{1},\ldots,i_{N},\hat{\bm{\xi}})	=\mathbb{P}\left(\bm{\xi}(\infty)=\sum_{a=1}^{N-1}\left(i_{a}\delta_{0}+(|\hat{\xi}_{a}|-i_{a})\delta_{L+1}\right)| \bm{\xi_{0}}=\hat{\bm{\xi}}\right)
	D(\bm{n},\bm{\hat{\xi}})= \frac{\prod_{x=1}^{m}n_{s_{x}}^{y_{x}}}{(\nu-m+1)}
Therefore, up to a constant, $\mathbb{E}_{\mu^{\text{stat}}}\left[D(\bm{n},\bm{\hat{\xi}})\right]$ gives the non-equilibrium stationary state $m-$points correlations among non-empty species of types $s_{i}\in\{1,\ldots,N-1\}$ situated at sites $y_{x}\in\{1,\ldots,L\}$ with $x\in \{1,\ldots,m\}$. This expectation can be computed by \eqref{ExptationSS} once the absorption probabilities is known. The issue is that these absorption probabilities fulfil some difference equations that are not easy to solve. For example, we report here the solution for the one points correlations. 
		\Delta_{x}^{L}\mathcal{P}(1,\delta_{s}^{y})=0\\
where $\Delta_{x}^{L}$ is the discrete Laplace operator on the chain of length $L$. The equations \eqref{gamblers-ruin} can be solved and using $\mathcal{P}(1,\delta_{s}^{y})=1-\mathcal{P}(0,\delta_{s}^{y})$ we obtain 
\textbf{Remark}: In case $|\alpha|=|\beta|=1$ we have 

>>>>

\newcommand{\changelocaltocdepth}[1]{%
  \addtocontents{toc}{\protect\setcounter{tocdepth}{#1}}%
  \setcounter{tocdepth}{#1}%
}
\newcommand{\id}{I}
\newcommand{\Xt}{\widetilde{X}}

@@

\begin{equation}\label{duality-expectation}

@@

As already pointed out in Subsection \ref{statementDualitySubsection}, the dual process voids the graph by piling up particles of species $\{1,\ldots,N-1\}$ in the extra sites and replacing them by types $N$. This property allows to characterize the steady state non-equilibrium distribution (assuming that it is unique) by the probability that dual particles are absorbed in the extra-sites, then called \textit{absorption probabilities}. Here we consider a chain of length $L$ where the end sites $1$ and $L$ are connected with two reservoirs. Each site can host $\nu\geq 1$ particles, chosen among the $N$ types. The process $\left(\bm{n}(t)\right)_{t\geq0}$ is defined on the space $\Omega=\bigtimes_{x=1}^{L} \Omega_{x}$ and the generator reads
We specialize the results of Subsection \ref{statementDualitySubsection} to this chain. The process is dual with respect to $(\bm{\xi}(t))_{t\geq0} $ defined on the enlarged configuration space
	\widetilde{\Omega}=N_{0}^{N}\times\bigtimes_{x=1}^{L} \Omega_{x}\times N_{0}^{N}
where $\mathcal{L}_{x,x+1}$ is given by \eqref{Generator} and where the boundaries are absorbing, i.e. 
		\widetilde{\mathcal{L}}_{left}f(\bm{\xi})&=|\alpha|\sum_{a=1}^{N-1}\xi_{a}^{1}\left(f(\bm{\xi}-\bm{\delta}_{a}^{1}+\bm{\delta}_{N}^{1}+\bm{\delta}_{a}^{0})f(\bm{\xi})\right)\\\widetilde{\mathcal{L}}_{right}f(\bm{\xi})&=|\beta|\sum_{a=1}^{N-1}\xi_{a}^{L}\left(f(\bm{\xi}-\bm{\delta}_{a}^{L}+\bm{\delta}_{N}^{L}+\bm{\delta}_{a}^{L+1})-f(\bm{\xi})\right)
where $0$ and $L+1$ are the extra-sites. The duality function reads
The dual process has absorbing boundaries, i.e. all dual particles are eventually absorbed at $0$ or at $L+1$ with a certain probability. This implies that the chain is eventually voided and the steady state correlations of the original process can be written via the absorption probabilities of the dual process as explained in the following. 
We call $\mu_{stat}$ the stationary measure of the the process $(\bm{n}(t))_{t\geq 0}$. We assume that it has an arbitrary configuration $\bm{n}$. For every $m\in\{1,\ldots,L\}$ we consider the dual configuration  $\bm{\widehat{\xi}}=\sum_{k=1}\delta_{s_{k}}^{y_{k}}$ with $y_{k}\in\{1,\ldots,L\}$ and $s_{k}\in \{1,\ldots,N-1\}$ for every $k\in \{1,\ldots,m\}$. The following  equalities holds thanks to the duality relation \eqref{duality-expectation}:
	\mathcal{P}(i_{1},\ldots,i_{N},\hat{\bm{\xi}})	=\mathbb{P}\left(\bm{\xi}(\infty)=\sum_{a=1}^{N-1}\left(i_{a}\delta_{0}+(|\hat{\xi}_{a}|-i_{a})\delta_{L+1}\right)| \bm{\xi}(0)=\hat{\bm{\xi}}\right)
	D(\bm{n},\bm{\hat{\xi}})= \frac{\prod_{k=1}^{m}n_{s_{k}}^{y_{k}}}{(\nu-m+1)}
Therefore, up to a constant, $\mathbb{E}_{\mu^{\text{stat}}}\left[D(\bm{n},\bm{\hat{\xi}})\right]$ gives the non-equilibrium stationary state $m-$points correlations among non-empty species of types $s_{k}\in\{1,\ldots,N-1\}$ situated at sites $y_{k}\in\{1,\ldots,L\}$ with $k\in \{1,\ldots,m\}$. This expectation can be computed by \eqref{ExptationSS} once the absorption probabilities are known. The issue is that these absorption probabilities fulfil some difference equations that are not easy to solve. For example, we report here the solution for the simplest case of one point correlations, since the single dual particle behaves as an independent random walker. 
\paragraph{One point correlations for the $\nu\geq 1$ case}
		\Delta_{y}^{L}\mathcal{P}(1,\delta_{s}^{y})=0\\
where $\Delta_{y}^{L}$ is the discrete Laplace operator on the chain of length $L$. The equations \eqref{gamblers-ruin} can be solved and using $\mathcal{P}(1,\delta_{s}^{y})=1-\mathcal{P}(0,\delta_{s}^{y})$ we obtain 
\textbf{Remark}: In case $|\alpha|=|\beta|=\nu=1$ (when the chain is integrable) we have 
that is the same result obtained in \cite{vanicat2017exact}. 
\\ Already for two-point correlations the computations become more involving and hard to solve. In the following we specialize the chain to the integrable case $\nu=1$. As it will be shown, in such a situation the non-equilibrium steady state correlations can be computed with different techniques exploiting the integrability of the process (matrix product ansatz and quantum inverse scattering method) combined with duality and, as a by-product, all absorption probabilities can be explicitly found. 
\subsection{Duality for the integrable process on a line segment}\label{integrableChain-duality}
Here we consider the chain defined in Subsection \ref{subsection-description-process-LINE}. We recall that the geometry is again a one dimensional chain with sites $\{1,\ldots,L\}$ where reservoirs are attached to $1$ and $L$, but with the hard-core exclusion rule, i.e. $\nu=1$. The dynamics is described by an Hamiltonian that is written in terms of the first fundamental representation of $gl(N)$. Duality result stated in Subsection \ref{statementDualitySubsection} can be adapted. The extra-sites are denoted by $0$ and $L+1$, and they are connected with sites $1$ and $L$ of the chain respectively. The matrix $R_{x}$ defined in \eqref{Revmatrix} 
reduces to the identity, therefore the relation between the generators of the algebra and its transposed reduces to

@@

Therefore, duality matrix reads

@@

and its elements are given by

@@

The dual Hamiltonian is
\begin{equation}\label{boundary-H-dual-chain}
	\widetilde{H}_{left}=\sum_{a=1}^{N-1}\left((a^{\dagger})_{a}^{0}\,E_{Na}^{1}-E_{aa}^{1}\right)\qquad 	\widetilde{H}_{right}=\sum_{a=1}^{N-1}\left((a^{\dagger})_{a}^{L+1}\,E_{Na}^{L}-E_{aa}^{L}\right)
In the long-time limit the dual process voids the chain, i.e. all particles of types $\{1,\ldots,N-1\}$ are eventually absorbed at the extra-sites $0$ and $L+1$ and replaced by types $N$. We also notice that, up the the bosonic creation operators $(a^{\dagger})_{a}$, the dual boundary Hamiltonians \eqref{boundary-H-dual-chain} are triangular. This simplification will be crucial in Sections \ref{sectionIntegrabiliy} and \ref{Section-Mapping}.\\

Using the above duality elements, we can compute the  $m$-point correlations between non-empty particles using the absorption probabilities. Consider a dual configuration $\bm{\hat{\xi}}:=\sum_{k=1}^{m}\delta_{s_{k}}^{y_{k}}$ where $s_{k}\in\{1,\ldots,N-1\}$ and $y_{k}\in\{1,\ldots,L\}$ such that $y_{k}<y_{k+1}$ for every $k\in\{1,\ldots,m\}$. Fix an arbitrary initial configuration $\bm{n}$, the elements \eqref{ElementsDualityMatrixChain} of the duality matrix computed in the dual configuration $\bm{\hat{\xi}}$ read
\begin{equation}
	D(\bm{n},\bm{\hat{\xi}})=\prod_{k=1}^{m}\mathbbm{1}_{\{n_{s_{k}}^{s_{k}}=1\}}
\end{equation}
Repeating the argument of Subsection \ref{Subsection-ss-nonI} and denoting the $m$-point correlations by $\langle n_{s_{1}}^{y_{1}}\cdots n_{s_{m}}^{y_{m}}\rangle_{stat}$ we have 
%To compute the steady state $m$-point correlations of the process $\bm{n}(t)$ we recall that the dual process eventually voids the chain (by putting all particles in the extra-sites $0$ and $L$). Therefore, we have the following equalities
\begin{equation}\label{CorrelationByABS-prob}
	\begin{split}
	\langle n_{s_{1}}^{y_{1}}\cdots n_{s_{m}}^{y_{m}}\rangle_{stat}&=\lim_{t\to \infty}\mathbb{E}_{\bm{n}}\left[D(\bm{n}(t),\bm{\hat{\xi}})\right]=\lim_{t\to \infty}\mathbb{E}_{\bm{\xi}}\left[D(\bm{n},\bm{\hat{\xi}}(t))\right]
	\\&=\sum_{c_{1}=0}^{1}\ldots \sum_{c_{m}=0}^{1}\prod_{k=1}^{m}\alpha_{a_{k}}^{c_{k}}\beta_{a_{k}}^{1-c_{k}}\mathcal{P}(c_{1},\ldots,c_{m},\hat{\bm{\xi}})
\end{split}
\end{equation}
where
\begin{equation}\label{absProbabilitiesIntegrable}
	\mathcal{P}(c_{1},\ldots,c_{m},\hat{\bm{\xi}}):=\mathbb{P}\left(\bm{\xi}(\infty)=\sum_{k=1}^{m}\left(c_{k}\delta_{0}+(1-c_{k})\delta_{L+1}\right)\lvert \bm{\xi}(0)=\hat{\bm{\xi}}\right)
\end{equation}
 namely the joint probability of having, for all $k=1,\ldots,m$, $c_{k}$ particles of species $s_{k}$ at site $y_{k}$ are absorbed at left and $1-c_{k}$ absorbed at right. Therefore, by computing these absorption probabilities we determine the arbitrary non-equilibrium steady state correlations of the integrable multi-species stirring process.

@@

In this section we aim to use duality and the matrix product ansatz to determine the non equilibrium steady state and its arbitrary correlation for the integrable version of the multi-species stirring process, defined in Subsection \ref{integrableChain-duality}. As it will turn out, duality simplifies the commutation relations of the matrix product ansatz allowing to determined a closed formula for all correlations. As a by-product, absorption probabilities can be explicitly determined, giving a further probabilistic interpretation for the non-equilibrium steady state. 

@@

This Markov chain is irreducible and positive recurrent, therefore there exists a unique stationary measure, that will be reached when time goes to infinity, regardless of the initial distribution. 

@@

The MPA gives an abstract form of the steady state in terms of an algebra of operators. However, to write explicitly the steady state one should provide a representation of this algebra. Another approach is to directly use the commutators and and the boundary actions compute correlations. This path has been followed in \cite{vanicat2017exact} to compute one-point and two-point correlations. 
Here we follow the same approach. We will show that the duality property established in Section \ref{sectionDuality} simplifies substantially the  use of commutators in the

@@

The first transformation $S_{1}$ is closely related to the dual matrix and we will see that $H^{'}$ is $\widetilde{H}^{T}$, up to the extra-site term described by the bosonic creation operator in \eqref{siteDualHamiltonian}. Therefore, $H^{'}$ is not stochastic. The second transformation $S_{2}$ diagonalize the right boundary. Finally, the composition  $S_{2} S_{1}$ allows to simplify the commutators of the MPA algebra. \\

The computation are organized as follows: we first define the similarity transformations $S_{1},S_{2}$ and the Hamiltonians $H^{'},H^{''}$. Second, we compute right eigenvector with vanishing eigenvalue of $H^{''}$ thanks to the simpler commutators. Finally, we invert the similarities and retrieve the right eigenvector with vanishing eigenvalue of $H^{'}$ and  the arbitrary correlations in the non-equilibrium steady state. The explicit expression of the steady state $|\Psi\rangle$ of $H$ is reported in Appendix \ref{appB}.

@@

\textbf{Remark:} the boundary Hamiltonians $H_{left}^{'}$ and $H_{right}^{'}$ relates to the transposed of  $\widetilde{H}_{left}$ and $\widetilde{H}_{right}$, i.e. to the transposed of the boundary part of the dual Hamiltonian  defined in \eqref{siteDualHamiltonian}. They are obtained replacing the  extra-site bosonic creation operators $(a^{\dagger})_{a}^{0}$  (resp. $(a^{\dagger})_{a}^{L+1}$) with the corresponding reservoir parameters $\alpha_a$ (resp. $\beta_a$).

@@

i.e. the left one is lower triangular and depends on the differences between boundary parameters, while the right one is diagonal. The expressions of $H_{left}^{''}$ and $H_{right}^{''}$ are proved using the Hadamard formula \eqref{HadamardFormula} with $B=\sum_{c=1}^{N-1}\beta_{c}e_{cN}$ and $A=e_{aN},$ and $A=e_{aa}$. We compute

@@

It is convenient to introduce also the whole transformation

@@

    \delta_{\tau_{x},N}:=\begin{cases}

@@

Using the transformed operators $(\widetilde{X}_{a})_{a\in\{1,\ldots,N\}}$, the vector $|\Psi^{''}\rangle$ of the matrix product ansatz is written as:

@@

By knowing the steady state  of the Hamiltonian $H^{''}$ we use \eqref{S2-Inverse} to retrieve the steady state of the Hamiltonian $H^{'}$. The result is the following 

@@
